---
title: "selective annotations"
format: html
author: Jen Richmond
date: 2025-05-23
code-fold: show
---

My friend Lisa called me with a ggplot challenge the other day. She was trying to work out how to have only some of her points labelled with annotations. Here are some notes about how you can label some points but not others. 

### load packages/make dataframe


```{r}
#| message: false
#| warning: false

library(tidyverse)

## make up some data

df <- data.frame(
             level = as.factor(c("Level A", "Level A", 
                                 "Level B", "Level B", 
                                 "Level C", "Level C", 
                                 "Level D", "Level D",
                                 "Level E", "Level E")),
              gender = as.factor(c("Female","Male",
                                   "Female","Male",
                                   "Female","Male",
                                   "Female","Male", 
                                   "Female","Male")),
             score = c(12L, 23L, 13L, 20L, 16L, 18L, 15L, 20L, 11L, 22L)
      )

glimpse(df)

```

# data setup 

First, check that levels are behaving and then create new variables that contain just the values you want to print on your plot. 

```{r}

levels(df$level)


df <- df %>%
  mutate(f_values = case_when(gender == "Female" ~ score), 
         m_values = case_when(gender == "Male" ~ score)) 

glimpse(df)
  
```


# basic plot

```{r}
#| message: false
#| warning: false

df %>%
  ggplot(aes(x = level, y = score, 
             colour = gender, group = gender)) +
  geom_point() + 
  geom_line() +
  scale_y_continuous(limits = c(0,30)) +
  theme_minimal()


```

# plot with selective annotations

Here we have added two `geom_text()` calls, one for m_values and one for f_values. Unfortunately the values end up on top of the points.  

```{r}
#| message: false
#| warning: false
df %>%
  ggplot(aes(x = level, y = score, 
             colour = gender, group = gender)) +
  geom_point() + 
  geom_line() +
  scale_y_continuous(limits = c(0,30)) +
  theme_minimal() +
  geom_text(aes(label = m_values), size = 3) +
  geom_text(aes(label = f_values), size = 3) 
```

Add a little vjust to shift the points up and down a little from the points. Don't love what has happened to the legend though...

```{r}
#| message: false
#| warning: false
df %>%
  ggplot(aes(x = level, y = score, 
             colour = gender, group = gender)) +
  geom_point() + 
  geom_line() +
  scale_y_continuous(limits = c(0,30)) +
  theme_minimal() +
  geom_text(aes(label = m_values), vjust = -2, size = 3) +
  geom_text(aes(label = f_values), vjust = 2, size = 3) 
```


Adding show.legend = FALSE to let R know that I don't want the text labels to appear in the legend. 


```{r}
#| message: false
#| warning: false
df %>%
  ggplot(aes(x = level, y = score, colour = gender, group = gender)) +
  geom_point() + 
  geom_line() +
  scale_y_continuous(limits = c(0,30)) +
  theme_minimal() +
  geom_text(aes(label = m_values), vjust = -2, size = 3, show.legend = FALSE) +
  geom_text(aes(label = f_values), vjust = 2, size = 3, show.legend = FALSE) 
```


SUCCESS!
